---
title: "Final Project QMEE"
author: "Alejandro Rojas"
output:
  html_document:
    fig_caption: yes
    fig_height: 10
    fig_width: 12
    keep_md: yes
---


File import into R maintaining the first row as header for the columns.  Also the samples from Arkansas 2011 were removed since this samples only came form a single field, not complying with the same sampling parameters used for the other samples.
```{r,echo=FALSE,message=FALSE}
setwd("~/Dropbox/Stats_Ecology-Evolution/Homeworks/Final_project/Final_Project_QMEE/")
library(ggplot2)
library(stargazer)
library(car)
library(arm)
library(effects)
require(MASS) # for glm.nb
require(bbmle)
require(emdbook) # for the dzin
require(lme4)
source("~/Dropbox//Stats_Ecology-Evolution/Homeworks/Hw-2/PartialR2_WP2012_ID.R")

oomy_data <- read.csv("data_qmee.csv", header = T)
oomy_data <- oomy_data[-1,]
```

```{r}
length(oomy_data$State)
stargazer(oomy_data, summary=TRUE, type="text")
```

Transform year data into factor and data type is checked after transformation.
```{r, echo=FALSE}
oomy_data$Year <- factor(oomy_data$Year)
oomy_data$ct.temp <- scale(x=oomy_data$Temp, center=TRUE, scale=FALSE)
oomy_data$ct.lat <- scale(x=oomy_data$Lat, center=TRUE, scale=FALSE)
oomy_data$ct.precp <- scale(x=oomy_data$Precipitation, center=TRUE, scale=FALSE)
```

Histogram representing the distribution of number of species found per year:
```{r, echo=FALSE}
#Histogram of number of species per year for the survey
ggplot(data=oomy_data, aes(x=OTU)) +
  geom_histogram(aes(fill=Year), binwidth=2, colour="black") + labs(x="OTU") + facet_grid(Year ~ .)
```

Box plots for number of species in each state, compiling the 6 fields per state
```{r}
#Boxplot for the number of species per state
ggplot(data=oomy_data, aes(y=OTU, x=State)) +
  geom_boxplot(aes(fill=Year)) + labs(y="OTU") + facet_grid(Year ~ .)
#Boxplot for the shannon diversity index per state
ggplot(data=oomy_data, aes(y=shannon, x=State)) +
  geom_boxplot(aes(fill=Year)) + labs(y="Shannon index") + facet_grid(Year ~ .)
```

Plotting the temperature against the number of species observed or using shannon diversity index, there was not a striking pattern that correlates the two factors.  
```{r, echo=FALSE}
#Evaluation of the relation of temperature and number of species across years
par(mfrow = c(1,1))
ggplot(data=oomy_data, aes(x=ct.temp, y=OTU)) + geom_point(colour="Black", size=4) +
  geom_point(aes(colour=Year), size=3) + stat_smooth(method=lm, aes(fill=Year)) + labs(x="Temperature (ºC)", y="OTU") 
#Evaluation of the relation of temperature and shannon index across years
ggplot(data=oomy_data, aes(x=ct.temp, y=shannon)) + geom_jitter(colour="Black", size=4) +
  geom_jitter(aes(colour=Year), size=3) + stat_smooth(method=lm, aes(fill=Year)) + labs(x="Temperature (ºC)", y="Shannon index") 
```

However, precipitation seems to have a more visible pattern plotting this against the number of species for both years.
```{r}
#Evaluation of the effect of precipitation on the number of species
ggplot(data=oomy_data, aes(x=ct.precp, y=OTU)) +
  geom_point(colour="Black", size=4) + geom_point(aes(colour=Year), size=3) + stat_smooth(method=lm, aes(fill=Year)) + labs(x="Precipitation (mm)", y="OTU")                                                                     

#Evaluation of the effect of precipitation on the Shannon diversity index
ggplot(data=oomy_data, aes(x=ct.precp, y=shannon)) + 
   geom_point(colour="Black", size=4) + geom_point(aes(colour=Year), size=3) + stat_smooth(method=lm, aes(fill=Year)) + labs(x="Precipitation (mm)", y="Shannon index")
```

Evaluation of latitude in relation to species diversity, in order to explore the relationships, and also test a ecological perspective.
```{r, echo=FALSE}
#Evaluation of the relation of latitude and number of species across years
par(mfrow = c(1,1))
ggplot(data=oomy_data, aes(x=ct.lat, y=OTU)) + geom_point(colour="Black", size=4) +
  geom_point(aes(colour=Year), size=3) + stat_smooth(method=lm, aes(fill=Year)) + labs(x="Latitude", y="OTU") 
#Evaluation of the relation of temperature and shannon index across years
ggplot(data=oomy_data, aes(x=ct.lat, y=shannon)) + geom_jitter(colour="Black", size=4) +
  geom_jitter(aes(colour=Year), size=3) + stat_smooth(method=lm, aes(fill=Year)) + labs(x="Latitude", y="Shannon index") 
```

As part of the model selection, different linear models were set, using diversity (Shannon index and OTU) as response variable, using temperature, latitude, precipitation as predictor variables. 
```{r}
reg.null <- lm(shannon ~ 1, data=oomy_data)
reg.lat <- lm(shannon ~ ct.lat, data=oomy_data)
reg.temp <- lm(shannon ~ ct.temp, data=oomy_data)
reg.prec <- lm(shannon ~ ct.precp, data=oomy_data)
reg.lat.temp <- lm(shannon ~ ct.lat + ct.temp, data=oomy_data)
reg.lat.prec <- lm(shannon ~ ct.lat + ct.precp, data=oomy_data)
reg.temp.prec <- lm(shannon ~ ct.precp + ct.precp, data=oomy_data)
reg.latXtemp <- lm(shannon ~ ct.lat*ct.temp, data=oomy_data)
reg.latXprec <- lm(shannon ~ ct.lat*ct.precp, data=oomy_data)
reg.tempXprec <- lm(shannon ~ ct.precp*ct.precp, data=oomy_data)

reg2.null <- lm(OTU ~ 1, data=oomy_data)
reg2.lat <- lm(OTU ~ ct.lat, data=oomy_data)
reg2.temp <- lm(OTU ~ ct.temp, data=oomy_data)
reg2.prec <- lm(OTU ~ ct.precp, data=oomy_data)
reg2.lat.temp <- lm(OTU ~ ct.lat + ct.temp, data=oomy_data)
reg2.lat.prec <- lm(OTU ~ ct.lat + ct.precp, data=oomy_data)
reg2.temp.prec <- lm(OTU ~ ct.precp + ct.precp, data=oomy_data)
reg2.latXtemp <- lm(OTU ~ ct.lat*ct.temp, data=oomy_data)
reg2.latXprec <- lm(OTU ~ ct.lat*ct.precp, data=oomy_data)
reg2.tempXprec <- lm(OTU ~ ct.precp*ct.precp, data=oomy_data)
```


```{r, echo=FALSE}
#Evaluation of all the models developed using Shannon as response
aic.shannon <- AICtab(reg.null,reg.lat,reg.temp,reg.prec,reg.lat.temp,reg.lat.prec,reg.temp.prec,reg.latXtemp,reg.latXprec,reg.tempXprec, weights=TRUE)

#Evaluation of all the models developed using OTUs as response
aic.OTU <- AICtab(reg2.null,reg2.lat,reg2.temp,reg2.prec,reg2.lat.temp,reg2.lat.prec,reg2.temp.prec,reg2.latXtemp,reg2.latXprec,reg2.tempXprec, weights=TRUE)
```

All the models were evaluated with delta AIC to determine the best fitting model using Shannon diversity index as a response variable.
```{r}
#AIC selection of model using OTU as response
aic.shannon
```

All the models were evaluated with delta AIC to determine the best fitting model using OTU number as a response variable.
```{r}
#AIC selection of model using OTU as response
aic.OTU
```

```{r, echo=FALSE}
#Selection of best models
models.shannon <- list(reg.lat,reg.lat.temp,reg.latXtemp,reg.latXprec) #Using shannon index
models.OTU <- list(reg2.latXprec,reg2.temp,reg2.lat.prec) #Using OTU

AIC.shannon <- AIC(reg.lat,reg.lat.temp,reg.latXtemp,reg.latXprec)
log.lik.shannon <- sapply(models.shannon,logLik,USE.NAMES=TRUE)

AIC.otu <- AIC(reg2.latXprec,reg2.temp,reg2.lat.prec)
log.lik.OTU <- lapply(models.OTU, logLik, USE.NAMES=TRUE)
```


The best fitting models were selected based on the delta AIC and other parameters were evaluated to confirm the selection of these models for both responses: Shannon diversity index and OTU number.
```{r}
#AIC, logLik and PRsq values for model using shannon
stargazer(AIC.shannon, type="text", summary = FALSE)
log.lik.shannon
sapply(models.shannon,PRsq)
#Summary for models using Shannon
stargazer(models.shannon, type="text", digits=3, omit.stat="f", omit.table.layout="n")

#AIC, logLik and PRsq values for models using OTU
stargazer(AIC.otu, type="text", summary = FALSE)
log.lik.OTU
sapply(models.OTU,PRsq)

#Summary for models usig OTU
stargazer(models.OTU, type="text", digits=3, omit.stat="f", omit.table.layout="n")
```

Evaluation of coefficient plots for shannon diversity index and OTU number.
```{r}
#Coefficient plots for Shannon diversity index
par(mfrow = c(2,2))
sapply(models.shannon,coefplot)

#Coefficient plots for OTU number
par(mfrow = c(2,2))
sapply(models.OTU,coefplot)
```

Evaluation of ACF on the residuals for the different models for both responses
```{r}
#ACF for models using Shannon
par(mfrow = c(2,2))
acf(resid(reg.lat))
acf(resid(reg.lat.temp))
acf(resid(reg.latXtemp))
acf(resid(reg.latXprec))

#ACF for models using Shannon
par(mfrow = c(2,2))
acf(resid(reg2.temp))
acf(resid(reg2.lat.prec))
acf(resid(reg2.latXprec))
```